---
title: "PS270 Final Project"
author: "Jonah Frumkin"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

  For this project, I will try to uncover disparities in the student achievement across race/ethnicity and household income. To do so, I will being providing data to answer my research question, which is, to what extent does race/ethnicity and median household income effect a student's SAT scores? The point of answering this question is to raise awareness to the inequities in the American education system and use the data provided in this project to help explain a phenomenon in American society. This phenomenon is the disparate impact happening in terms of socioeconomic inequality and mobility across racial/ethnic groups in the US. While the data I provide won't necessarily draw a causation between educational inequality and racial inequality, it will help draw an inference to get people thinking about the relationship between the two.

  I believe an increased household income will have a greater positive impact on white students' SAT scores compared to Black and Hispanic students because of disproportionate access to social capital and due to racial implicit bias. One would assume that with a higher median household income, a White student and a Black student would progress in a very similar manner over time due to increased financial power to send their child to a better school and provide access to extra support such as therapy and tutoring. However, I believe that holding wealth constant, White students will still perform better academically than Black students academically because financial disparities aren't the only factor at play here as there are more complex social factors at hand.
  
## Data Section

  One of my independent variables required me to analyze a data set from the US Census Bureau, which provides the median household income by zip code in New York City in 2015. The US Census Bureau collects this information through surveys such as the American Community Survey (ACS). Another one of my independent variables is the NYC Department of Education that report the race/ethnicity demographics of schools in NYC through a self-reported form, which the NYC DOE aggregates for each school. Finally, my dependent variable is the average SAT score of each public school in NYC, which is recorded by the College Board and then aggregated for each school by the NYC DOE. The data that the NYC DOE is aggregating comes from a data set in Kaggle for the 2014-2015 school year which I merged with a data set from the US Census Bureau, reporting median income household by zip code from 2015.  
  
  One of my independent variables, the race/ethnicity of a school, will separated into two categories "Majority White" if the student body is 50% white or over 50% white and "Minority White" if otherwise. My other independent variable, median household income by zip code (US Census Bureau data set) is determined by a five digit code that I will merge with the zip code of each school reported in the Kaggle data set. An observed pattern that would prove my hypothesis to be true is if schools with higher percentages of white students are found to have relatively higher SAT scores. Another pattern that would prove my hypothesis to be true is if there were distinct variations in SAT scores between schools with different median household incomes. Also, if there is a greater variation between SAT scores for white students across income groups compared non-white students, which would show the unfair racial biases that promote better outcomes for white students across all income groups. An observed pattern that would disprove this hypothesis would be if there was no general distinction in SAT scores between schools that had different proportions of white students. Also if there was not variation between income groups and if race/ethnicity ended up having no effect on SAT scores across different median household income groups. 

## Milestone 3

```{r}
library(tidyverse)
library(knitr)
nycsat <- read_csv("data/scores.csv")
nycincome <- read.csv("data/Median2015.csv")
```
```{r}
nycincome_clean <- nycincome |>
  slice(-1:-3) |>
  rename(`Zip Code` = NAME, `Household Median Income` = S1903_C02_001E) |>
  mutate(`Zip Code` = str_remove(`Zip Code`, "ZCTA5 ")) |>
  mutate(`Zip Code` = as.numeric(`Zip Code`)) |>
  mutate(`Household Median Income` = as.numeric(`Household Median Income`)) |>
    select(`Zip Code`, `Household Median Income`) 

  
```

```{r}
nycsat_clean <- nycsat |>
  mutate(`Average SAT Score` = `Average Score (SAT Math)` + `Average Score (SAT Reading)` + `Average Score (SAT Writing)`) |>
  filter( `Percent Tested` >= "50%") |>
  rename(`Standard Percent Tested` = `Percent Tested`) |> 
  mutate(`Percent White` = str_remove(`Percent White`, "%")) |>
  mutate(`Percent White` = as.numeric(`Percent White`)) |>
  select(`School ID`, `Zip Code`, `Average SAT Score`, `Percent White`, `Student Enrollment`, `Standard Percent Tested`)
```


```{r}
nycsat_income <- nycsat_clean |>
  left_join(nycincome_clean, by = "Zip Code") |>
  drop_na(`Household Median Income`)
  
  
  
```


```{r}
income_group_plot <- ggplot(nycsat_income, aes(x = `Household Median Income`, y = `Average SAT Score`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "NYC Public School SAT Scores (2014-15) by Median Household Income", caption = "Source: US Census Bureau")

income_group_plot

Model_income <- lm(`Average SAT Score` ~ `Household Median Income`, nycsat_income)
Model_income


```
```{r}
race_eth_plot <- ggplot(nycsat_income, aes(x = `Percent White`, y = `Average SAT Score`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "NYC Public School SAT Scores (2014-15) by Race/Ethnicity", caption = "Source: New York City Department of Education and College Board") 

race_eth_plot

Model_race_eth <- lm(`Average SAT Score` ~ `Percent White`, nycsat_income)
Model_race_eth
```




## Milestone 4


```{r}
nycincome_clean_2 <- nycincome |>
  slice(-1:-3) |>
  rename(`Zip Code` = NAME, `Household Median Income` = S1903_C02_001E) |>
  mutate(`Zip Code` = str_remove(`Zip Code`, "ZCTA5 ")) |>
  mutate(`Zip Code` = as.numeric(`Zip Code`)) |>
  mutate(`Household Median Income` = as.numeric(`Household Median Income`)) |>
   select(`Zip Code`, `Household Median Income`) 

nycsat_clean <- nycsat |>
  mutate(`Average SAT Score` = `Average Score (SAT Math)` + `Average Score (SAT Reading)` + `Average Score (SAT Writing)`) |>
  filter( `Percent Tested` >= "50%") |>
  rename(`Standard Percent Tested` = `Percent Tested`) |>
  mutate(`Majority White` = if_else(`Percent White` > "50%", "Majority White", "Minority White")) |>
  select(`School ID`, `Zip Code`, `Average SAT Score`, `Majority White`, `Student Enrollment`, `Standard Percent Tested`)

nycsat_income_2 <- nycsat_clean |>
  left_join(nycincome_clean_2, by = "Zip Code") |>
  drop_na(`Household Median Income`)
  

```
```{r}
nycsat_race_eth <- nycsat_income_2 |>
  group_by(`Majority White`) |>
  summarise(`Median SAT Score` = median(`Average SAT Score`, na.rm = TRUE)) 

kable(nycsat_race_eth, digits = 0, col.names = c("Race/Ethnicity", "Median SAT Score"), caption = "Median SAT Score by Race/Ethnicity")
  
```


```{r}
scatter_nycsatinc_plot <- 
  ggplot(nycsat_income_2, aes(x = `Household Median Income`, y = `Average SAT Score`, color = `Majority White`)) +
  geom_point() +
  geom_smooth(aes(group = `Majority White`), method = "lm", se = FALSE) +
  labs(title = "Effect of Household Median Incomes on SAT Scores by Race/Ethnicity",  caption = "Sources: New York City Department of Education, College Board, and US Census Bureau")

scatter_nycsatinc_plot  

model_majority <- lm(`Average SAT Score` ~ `Household Median Income`, data = subset(nycsat_income_2, `Majority White` == "Majority White"))
model_minority <- lm(`Average SAT Score` ~ `Household Median Income`, data = subset(nycsat_income_2, `Majority White` == "Minority White"))

model_majority
model_minority
```


#Description of Data

My hypothesis was upheld by the data presented by the NYC Department of Education, College Board and US Census Bureau from the 2014-15 school year. I joined two data sets by the zip codes in NYC so I was able to analyze the effect of race/ethnicity on SAT scores, median household income on SAT scores, and the extent to which race/ethnicity effects SAT scores across median household incomes. There was a positive relationship between percentage of white students and median household income on SAT scores. However, it was surprising to see that race/ethnicity had a huge effect on SAT scores relative to household income. Although there was a lot more data for minority race/ethnic schools compared to data for schools with white students, making the data subject to error, this data upholds the consensus that systemic disparities occur throughout the American education system because of historic racist policies and practices.  



