---
title: "PS270 Final Project"
author: "Jonah Frumkin"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

  For this project, I will try to uncover disparities in student achievement across race/ethnicity and household income. To do so, I will be analyzing data from the New York City Department of Education, College Board, and US Census Bureau to answer my research question, which is, to what extent do race/ethnicity and median household income affect a student's SAT scores? The point of answering this question is to provide evidence-based research to show the inequity in the American education system. This can ultimately provide grounding for assumptions that broader systemic disparities exist in the US such as the disparate impact happening in terms of socioeconomic inequality and mobility between racial/ethnic groups. While the data I provide won't necessarily draw a causation between educational inequality and racial inequality, hopefully, it can get people to question why these disparities exist.

  Before analyzing the data, I hypothesized that an increased household income would have a greater positive impact on white students' SAT scores compared to Black and Hispanic students because of disproportionate access to symbolic advantages that come from being white such as social capital and white privilege. In a world without race as a defining factor of identity, one would assume that with a higher median household income, a White student and a Black student would progress in a very similar manner over time due to increased financial power to send their child to a better school and provide access to extra support such as therapy and tutoring. However, I believe that holding wealth constant, White students will still perform better academically than Black students academically because financial disparities aren't the only factor at play.

  
  
  
## Data Section

```{r}
library(tidyverse)
library(knitr)
library(broom)
library(modelsummary)
nycsat <- read_csv("data/scores.csv")
nycincome <- read.csv("data/Median2015.csv")
```
# Data from US Census Bureau 

```{r}
nycincome_clean <- nycincome |>
  slice(-1:-3) |>
  rename(`Zip Code` = NAME, `Household Median Income` = S1903_C02_001E) |>
  mutate(`Zip Code` = str_remove(`Zip Code`, "ZCTA5 ")) |>
  mutate(`Zip Code` = as.numeric(`Zip Code`)) |>
  mutate(`Household Median Income` = as.numeric(`Household Median Income`)) |>
    select(`Zip Code`, `Household Median Income`) 

  
```
# Data from Kaggle (NYC Department of Education and College Board)
```{r}
nycsat_clean <- nycsat |>
  mutate(`Average SAT Score` = `Average Score (SAT Math)` + `Average Score (SAT Reading)` + `Average Score (SAT Writing)`) |>
  filter( `Percent Tested` >= "50%") |>
  rename(`Standard Percent Tested` = `Percent Tested`) |> 
  mutate(`Percent White` = str_remove(`Percent White`, "%")) |>
  mutate(`Percent White` = as.numeric(`Percent White`)) |>
  mutate(`Standard Percent Tested` = str_remove(`Standard Percent Tested`, "%")) |>
  mutate(`Standard Percent Tested` = as.numeric(`Standard Percent Tested`)) |>
  select(`School ID`, `Zip Code`, `Average SAT Score`, `Percent White`, `Student Enrollment`, `Standard Percent Tested`)
```

# Merging data sets
```{r}
nycsat_income <- nycsat_clean |>
  left_join(nycincome_clean, by = "Zip Code") |>
  drop_na(`Household Median Income`)
```
# Outcome variable

```{r}
dep_nycsat_income_plot <-
  ggplot(nycsat_income, aes(x = `Average SAT Score`)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 100) +
  labs(title = "Proportion of Average SAT Scores in NYC Public Schools 2014-15", caption = "Source: New York City Department of Education and College Board" )

dep_nycsat_income_plot
```

  One of my independent variables required me to analyze a data set from the US Census Bureau, which provides median household income by zip code in New York City in 2015. The US Census Bureau collects this information through surveys such as the American Community Survey (ACS). Another one of my independent variables is collected using data from the NYC Department of Education that reports the race/ethnicity demographics of schools in NYC through a self-reported form, which the NYC DOE aggregates for each school. The four categories of race/ethnicity in this data set are White, Black, Hispanic, and Asian. My dependent variable is the average SAT score of each public school in NYC, which is recorded by the College Board and then aggregated for each school by the NYC DOE. The data that the NYC DOE is aggregating (SAT scores and race) comes from a data set in Kaggle for the 2014-2015 school year which I merged with a data set from the US Census Bureau, reporting median income household by zip code from 2015.
  
  One of my independent variables, the race/ethnicity of a school, will first be a quantitative variable stating the percentage white students in a school and then will be converted into a categorical variable with two categories, “Majority White” if the student body is 50% white or over 50% white and “Minority White” if otherwise. My other independent variable, median household income by zip code (US Census Bureau data set) is determined by a five-digit code that I will merge with the zip code of each school reported in the Kaggle data set. 
  
  The main visualization I will be analyzing for this study is Vizualization 4 in the results section, which shows the difference between the effect of a majority white school on SAT scores vs a minority white school. Therefore, this research design is cross-sectional because it is comparing the the treatment effect of my tested independent variables at a single point in time. 
  
  The “dep_nycsat_income_plot” shows the density of average SAT scores of NYC public schools. The data shows that without taking race/ethnicity and median household income by zip code into account, a plurality of schools have average SAT scores from 1200-1400 since most of the data is concentrated in that range.
  
  

    
## Results section

# Vizualization 1 
```{r}
income_group_plot <- ggplot(nycsat_income, aes(x = `Household Median Income`, y = `Average SAT Score`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "NYC Public School SAT Scores (2014-15) by Median Household Income", caption = "Source: US Census Bureau")

income_group_plot

Model_income <- lm(`Average SAT Score` ~ `Household Median Income`, nycsat_income)


tidy(Model_income) |>
   kable(caption = "Regression Results for Household Median Income")

Model_income_r_squared <- summary(Model_income)$r.squared
cat("The R-squared value for Household Median Income is:", round(Model_income_r_squared, 3), "\n\n")

Model_income_race_control <- lm(`Average SAT Score` ~ `Household Median Income` + `Percent White`, nycsat_income)

tidy(Model_income_race_control) |>
   kable(caption = "Regression Results for Household Median Income when controlling for race/ethnicity")

Model_income_race_control_r_squared <- summary(Model_income_race_control)$r.squared
cat("The R-squared value of Household Median Income when controlling for race/ethnicity is:", round(Model_income_race_control_r_squared, 3), "\n\n")


inc_sat_cor <- cor(nycsat_income$`Average SAT Score`, nycsat_income$`Household Median Income`)
cat("The correlation between Household Median Income and Average SAT Score is:", round(inc_sat_cor, 3))



```
# Vizualization 2

```{r}
race_eth_plot <- ggplot(nycsat_income, aes(x = `Percent White`, y = `Average SAT Score`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "NYC Public School SAT Scores (2014-15) by Race/Ethnicity", caption = "Source: New York City Department of Education and College Board") 

race_eth_plot

Model_race_eth <- lm(`Average SAT Score` ~ `Percent White`, nycsat_income)


tidy(Model_race_eth) |>
   kable(caption = "Regression Results for race/ethnicity")

Model_race_eth_r_squared <- summary(Model_race_eth)$r.squared
cat("The R-squared value for race/ethnicty is:", round(Model_race_eth_r_squared, 3), "\n\n")

race_sat_cor <- cor(nycsat_income$`Average SAT Score`, nycsat_income$`Percent White`)
cat("The correlation between race/ethnicity and Average SAT Score is:", round(race_sat_cor, 3))

```
# Mutating Race/Ethnicity Variable
```{r}
nycincome_clean_2 <- nycincome |>
  slice(-1:-3) |>
  rename(`Zip Code` = NAME, `Household Median Income` = S1903_C02_001E) |>
  mutate(`Zip Code` = str_remove(`Zip Code`, "ZCTA5 ")) |>
  mutate(`Zip Code` = as.numeric(`Zip Code`)) |>
  mutate(`Household Median Income` = as.numeric(`Household Median Income`)) |>
   select(`Zip Code`, `Household Median Income`) 

nycsat_clean <- nycsat |>
  mutate(`Average SAT Score` = `Average Score (SAT Math)` + `Average Score (SAT Reading)` + `Average Score (SAT Writing)`) |>
  filter( `Percent Tested` >= "50%") |>
  rename(`Standard Percent Tested` = `Percent Tested`) |>
  mutate(`Majority White` = if_else(`Percent White` > "50%", "Majority White", "Minority White")) |>
  select(`School ID`, `Zip Code`, `Average SAT Score`, `Majority White`, `Student Enrollment`, `Standard Percent Tested`)

nycsat_income_2 <- nycsat_clean |>
  left_join(nycincome_clean_2, by = "Zip Code") |>
  drop_na(`Household Median Income`)
  

```
# Vizualization 3 

```{r}
nycsat_race_eth <- nycsat_income_2 |>
  group_by(`Majority White`) |>
  summarise(`Median SAT Score` = median(`Average SAT Score`, na.rm = TRUE)) 

kable(nycsat_race_eth, digits = 0, col.names = c("Race/Ethnicity", "Median SAT Score"), caption = "Median SAT Score by Race/Ethnicity")
  
```

# Vizualization 4

```{r}
scatter_nycsatinc_plot <- 
  ggplot(nycsat_income_2, aes(x = `Household Median Income`, y = `Average SAT Score`, color = `Majority White`)) +
  geom_point() +
  geom_smooth(aes(group = `Majority White`), method = "lm", se = FALSE) +
  labs(title = "Effect of Household Median Incomes on SAT Scores by Race/Ethnicity",  caption = "Sources: New York City Department of Education, College Board, and US Census Bureau")

scatter_nycsatinc_plot  

model_majority <- lm(`Average SAT Score` ~ `Household Median Income`, data = subset(nycsat_income_2, `Majority White` == "Majority White"))
model_minority <- lm(`Average SAT Score` ~ `Household Median Income`, data = subset(nycsat_income_2, `Majority White` == "Minority White"))


tidy(model_majority) |>
  kable(caption = "Regression Results for Majority White Schools")
tidy(model_minority) |>
  kable(caption = "Regression Results for Minority White Schools")


```

  In the first of 4 visualizations of the results section, the “income_group_plot” shows a positive relationship between median household income and average SAT scores of NYC public schools. The main coefficient of interest here is median household income and according to the plot, average SAT scores increased by 1.35 points for an additional 1,000 dollars in median household income. However, the regression model tied to this visualization shows that when controlling for race/ethnicity SAT scores only increased by 0.25 points for an additional 1,000 dollars showing that the racial composition of a student body has an extremely large effect on SAT scores. Furthermore, the R-squared for this coefficient is only about 0.03 and the correlation is about 0.19 showing that income may not be the greatest indicator of student achievement. 
  
  In the second visualization, the “race_eth_plot” a positive relationship is shown between race/ethnicity and average SAT scores of NYC public schools. The main coefficient of interest here is race/ethnicity and according to the plot, average SAT scores increase by 83 points when a school’s composition of white students increases by 10 percentage points. This is extremely statistically significant because it shows that the race/ethnicity of a school has a direct effect on average SAT scores. For example, schools with a higher proportion of white students compared to students of color, are more likely to have a higher average SAT score. In addition, this data is statistically significant because race/ethnicity is shown to have notable implications for SAT scores as the R-squared for this regression model is about 0.4 and the correlation between these two variables is about 0.61. While there are other factors at play here, race/ethnicity is proving to be one of the main indicators of student achievement. 
  
  After mutating the race/ethnicity variable from a percentage to a binomial distribution of “majority white” and “minority white,” I display a third visualization. This confirms the presence of racial disparities in education as the “nycsat_race_eth” table shows the median of average SAT scores from schools with a majority of white students (1330) is 77 points higher than schools with a minority of white students (1253). 
  
  However, the fourth visualization is the main plot that I want to focus on because it defeats the myth that students of a different race/ethnicity can succeed at the same rate when their family works hard to increase their median household income. In this plot, the main coefficient of interest is median household income, but this coefficient is compared across racial/ethnic lines. Average SAT scores increased by 1.37 points for an additional 1,000 dollars in median household income for schools with a majority of white students while average SAT scores increased by 1.15 points for an additional 1,000 dollars in median household income for schools with a minority of white students. This is statistically significant because this shows that a school with a majority of white students will have on average a 22-point advantage in SAT scores over a school with a majority of white students when median household income increases by 100,000 dollars.


## Conclusion Section

# Testing for Confounders
   
```{r}

model_standard_percent_tested <- lm(`Average SAT Score` ~ `Standard Percent Tested`, nycsat_income)

model_standard_percent_tested_r_squared <- summary(model_standard_percent_tested)$r.squared
cat("The R-squared value for schools with 50% or more students tested is:", round(model_standard_percent_tested_r_squared, 3), "\n\n") 

race_eth_perc_con <- lm(`Average SAT Score` ~ `Percent White` + `Standard Percent Tested`, nycsat_income)

race_eth_perc_con_r_squared <- summary(race_eth_perc_con)$r.squared
cat("The R-squared value for race/ethnicity when controlling for schools with 50% or more students tested is:", race_eth_perc_con_r_squared, 3)

```

```{r}
stan_test <- cor(nycsat_income$`Average SAT Score`, nycsat_income$`Standard Percent Tested`)
cat("The correlation between Average SAT Score and Standard Percent Tested is:", round(stan_test, 3))
```

 My hypothesis was upheld by the data presented by the NYC Department of Education, College Board, and US Census Bureau from the 2014-15 school year. There was a greater positive relationship between median household income on SAT scores among majority white schools compared to minority white schools. when analyzing median household income as the main coefficient, there were relatively low implications for a school's median SAT score compared to other coefficients such as race/ethnicity and percentage of students tested. For example, when the coefficient of determination is race/ethnicity about 40% of the variation in SAT scores can be attributed to either of these independent variables while median household income could only explain 3%. This can help defeat the notion that the “American Dream” applies to all people no matter their race/ethnicity equally. This can be concluded because SAT scores play a large role in college admissions and if race/ethnicity is playing a much larger role in SAT scores than income, that means race/ethnicity can play a large role in socioeconomic mobility since a college degree has proven to help in upward mobility. 

Limitations to my analysis of race/ethnicity on student success can be attributed to the fact that this study only takes data from NYC public schools. For example, the data set I analyze only takes into account 277 schools when there are around 100,000 total public schools in the US. As a result, the location of the schools in my study could skew the data as the American education system is heavily influenced by different states. As a result, the outcome of students in one state could differ drastically from that of another state due to state policies, funding, and structure for schools. In addition, there are only 32 majority white schools compared to 245 minority white schools, which shows how there could be a lot of error after analyzing this binomial distribution because they are extremely uneven. Other data that could potentially alter the effect of median household income on SAT scores is if parents send their kids to school in a different zip code from which they live. 

Another limitation of this analysis can be due to confounding variables such as the “Standard Percent Tested” variable, which I created by filtering out schools that didn’t have over 50% tested students. The R-squared value for this variable is about 0.4, which means about 40% of the variation in SAT scores can be attributed to this variable. In addition, there is a strong positive relationship between this variable and SAT scores as it was 0.61. While the “Standard Percent Tested” variable wasn’t one of the variables I was analyzing to describe disparities in SAT scores, it proved to be a significant indicator of SAT scores. 

If I had more time, I would analyze a larger variety of school districts because that would be more representative of the broader US public school population. In addition, I would take this data from multiple school years so that I would be able to see if disparities among different groups of students have changed at all. Finally, with more schools being analyzed, I would be able to increase the “Standard Percent Tested” threshold to 80% or higher so that it wouldn’t be such a confounding variable anymore.
